---
title: "Intro to Stats - 0: R basics"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Variables
### Exercise 0 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Creating variables
To attribute a value to a variable, we use the <- operator, like below.  


```{r variables, exercise=TRUE, exercise.eval=TRUE}
#creating variables
a <- 1
print(a)


b <- sqrt(5)
b


```

A couple of important things here:  
- print() is a *function*. Functions are usually a verb followed by an argument (or more) between parenthesis: do_this(to_this); do_this(to_this, to_that, with_this).  
Can you find another function in that chunk of code?   

#### Exercise 
Add a line in the code above to create a new variable, c, that is the sum of a and b.

```{r variables-hint}

c <- a+b
```

### Text variables
A variable can also be text: then it has to be between " ". Add code to print the following variable 
```{r hei, exercise=TRUE, exercise.eval=TRUE}
text1 <- "Hei"


```

## Numerical vectors
### Creating vectors
Vectors are created by concatenating numbers in a list. The function c() concatenates elements that are listed between commas like so:
```{r vector, echo = TRUE}
v <- c(1,36,42)
v
```

You can also create a vector of consecutive numbers like so:

```{r vector2, echo=TRUE}
v2<- c(1:5)
v2
```

You can use "length" to know the length of your vectors:
```{r length, exercise=TRUE}
length(v)
length(v2)
```

```{r question1}
  question("What is 'length'?",
    answer("a vector"),
    answer("a list"),
    answer("a function", correct = TRUE),
    answer("a variable")
  )
```

### Combining vectors
Vectors can also be combined:

```{r combining-vectors, exercise=TRUE, exercise.eval=TRUE}
#combining vectors
v3 <- c(v,v2)
v3
```

This can be done recursively. Try to combine v3 to v:
```{r combining-vectors-2, exercise=TRUE}

```

You can also vectors of different types, like numbers and text:

```{r mixed-vector, exercise=TRUE, exercise.eval=TRUE}
mix <- c(v3, text1)
mix
```


BUT: the numbers are now understood as character strings so you can't use them for calculation. 

### Structures
To know what is the type of a vector, you can use the "structure" function: str()
```{r str, exercise=TRUE}
str(v)

str(mix)
```
The first vector contains 3 _num_ = numerical elements. The second vector contains 9 _chr_ = character strings.  

### Manipulating structures
How do I change the type of a vector?
```{r str-manip, echo=TRUE}
v.char <- as.character(v)
v.char
str(v.char)
```

Now try to change the structure of v2:
```{r str-manip-ex, exercise=TRUE}

```

### Truth values
Use "==" to check if two items are equal. Use "=" or "<-" to attribute the value to something (it's best to stick to "<-" to attribute values so you don't make mistakes.)

```{r equals, echo=TRUE}
v==36
```

```{r equal-demo, echo=TRUE}
d = 3
d
d == 3
```

You can also use superior to: >, inferior to: <, or different: !=, to compare or find values. You can use the *which* function to find certain numbers inside a vector:
```{r which, echo=TRUE}
which(v<40)
```
This tells you where in the vector you can find numbers that are inferior to 40 (it gives their index). Now you can also find which numbers those are:

```{r which2, echo=TRUE}
v[which(v<40)]
```

Try to find if there are numbers superior to 30 in v3 and which ones.
```{r which-exo, exercise=TRUE, exercise.lines=3}

```

## Factors
Let's imagine we have 13 dogs of three different colours.
```{r factor-creation, exercise=TRUE, exercise.eval=TRUE}
dogCol <- c(rep("black", 5), rep("bown",5), rep("white",3))
dogCol <- factor(dogCol)
dogCol
```
You will notice that we used the *factor* function to change the vector into a factor. What was dogCol before it was changed into a factor?

```{r factor-creation-hint}
#use the str() function before changing dogCol into a factor.
```

### Levels
We can check the levels of our factor with the *levels* function:

```{r levels, exercise=TRUE}
levels(dogCol)
```

Levels are always alphabetically ordered. If you want a different one as the first level, use the *relevel* function with, as arguments, (1) the factor whose levels you want to reorder, and (2) the level you want as the first level:

```{r relevel, exercise=TRUE}
relevel(dogCol, "white")
```

You can try modifying the code to relevel differently.

### Modifying levels

Oops, there is a mistake: it should be "brown", not "bown". How do I change it?
```{r check-level, exercise=TRUE}
#first, check the index of the level you want to change
levels(dogCol)
```
You can see that the order of the levels is the same as the first one, alphabetical. If we had wanted to save the reordered level, how would we have done it?

```{r save-relevel, exercise=TRUE}

```


```{r save-relevel-hint}
dogCol<- relevel(dogCol, "white")
```

The index of "bown" is 2. This is how we can change it's label:

```{r relabel, exercise=TRUE}
levels(dogCol)[2]<- "brown"
#check
levels(dogCol)
```

## Loops and conditional statements

The following is a loop.This iterates over all the items in dogCol (i). It checks if the item *i* is between 1 and 3 (i.e., it will only do something for the first 3 items in the vector). If it is so, it replaces the value of that item in dogCol by a the value "white".
```{r loop, exercise=TRUE, exercise.eval=TRUE}
for (i in 1:3){
  dogCol[i]<- "white"
}
dogCol
#you define the "i" between ()
#the consequence is between {}
```

This next snippet is an IF/conditional statement. It will check whether a condition is matched and act accordingly. The structure is the following:   
  if (a logical expression testing a condition){   
  what to do if it is true  
  }else if (another logical expression){   
  what to do then   
  }else {what to do if all expressions above are FALSE}  

```{r if, exercise=TRUE}
if (length(v)==3){
  print("V has a length of 3.")
} else {
  print("V does not have a length of 3.")
}
```

You can combine both statements. For example, you can iterate over all items of a list / vector and do something different depending on the value you find. 

```{r for-if, exercise=TRUE}
#ex: let's iterate over all the items in dogCol, so i has to be between 1 (the index of the first item) and the index of the last item of the list, which corresponds to the number of items there are in the list.
for (i in 1:length(dogCol)){
  if (dogCol[i]=="white"){ #this checks if the item we're looking at right now is "grey".
  cat("The dog is ", as.character(dogCol[i]), "\n") #this will print the first part followed by the item which has i as its index. So if i=1 it will print the first item in dogCol, etc.
} else {cat("The dog is not white but it is ", as.character(dogCol[i]), "\n")}
}

```

## Data frames
A data frame is a matrix with rows (observations) and columns (variables). Let's take our dogs again:

```{r dogs, echo=TRUE}
dogs <- c(1:13)
dogsCol <- c(rep("black", 5), rep("brown",5), rep("white",3))
```

We have two vectors. We're going to convert them into a data frame with the data.frame() function. This enables us to rename our variables:
```{r dogSet1, echo=TRUE}
niceDogs <- data.frame(dogID = dogs, dogColour = dogsCol)
```

There's another way to do it:
```{r dogSet2, echo=TRUE}
niceDogs2<-as.data.frame(cbind(dogs, dogsCol))
```

Have a look at the first lines of both to see what they look like:
```{r headdogs, exercise=TRUE}

```


Some useful functions for data frames include *cbind()* to add a new column to our dataset or *rbind()* to add a new row.  
Both work the same way: cbind(dataframe, new column). (But don't forget to attribute it to a new variable if you want to save it!).  
Here is a vector with the dog's age. Add it as a new column to the niceDogs dataset and save that as niceDogs3.
```{r cbind, exercise=TRUE}
ageDogs <- c(3,2,5,13,1,4,9,11,6,12,5,7,8)

```

```{r cbind-hint}
niceDogs3<-cbind(niceDogs, ageDogs)
```

Now add the following dog to niceDogs
```{r doggy14, exercise=TRUE}
doggy14 <- data.frame(dogID=14, dogColour="grey")


```

```{r doggy14-hint}
niceDogs4<-rbind(niceDogs, doggy14)
```

You can also figure out the size of your data frame with nrow() and ncol()
```{r size, exercise=TRUE}
nrow(niceDogs)
ncol(niceDogs)
```

